---
title: 'Homework: iterative method'
author: "Ruzhang Zhao"
output:
  html_document:
    df_print: paged
---

```{r setup, include = FALSE}
source(file.path("..", "R", "util.R"))
required_packages <- c('RSpectra','tidyverse','testit')
install_and_load_packages(required_packages)
```

# Problem 1
Auto regressive processes can be viewed as a discrete analog of Ornstein–Uhlenbeck process &mdash; which coincides with Gaussian process based on an exponential covariance matrix &mdash; and hence is an example of Gaussian Markov random fields.
For instance, stationary lag-1 auto-regressive process 
$$x_t = \phi x_{t - 1} + \sqrt{1 - \phi^2} \, \epsilon_t, 
  \quad x_0 \sim \mathcal{N}(0, 1), 
  \quad \epsilon_t \mathbin{\overset{\small \textrm{i.i.d.}}{\sim}} \mathcal{N}(0, 1)$$
has the _tri-diagonal_ precision matrix
$$\boldsymbol{\Sigma}^{-1} = \frac{1}{1 - \phi^2} 
  \begin{bmatrix} 
  1 & -\phi & 0 & & & \ldots & 0 \\
  -\phi & 1 + \phi^2 & -\phi & 0 & & & \vdots \\
  0 & -\phi & 1 + \phi^2 & -\phi & 0 & & \\
    & & \ddots & \ddots & \ddots & & \\
    & & 0 & -\phi & 1 + \phi^2 & -\phi & 0 \\
  \vdots & &   & 0 & -\phi & 1 + \phi^2 & -\phi \\
  0 & \ldots &   &   & 0 & -\phi & 1\\
  \end{bmatrix}.$$
More generally, a lag-$k$ (non-stationary) auto-regressive process has a _banded_ precision matrix with bandwidth $k$.

Implement a fast matrix-vector $\boldsymbol{v} \to \boldsymbol{\Sigma}^{-1} \boldsymbol{v}$ operation, exploiting the structure of the AR-1 precision matrix.
Then use this function to find the top 10 principal components of $\boldsymbol{\Sigma}$ (not $\boldsymbol{\Sigma}^{-1}$) via Lanczos algorithm provided via `RSpectra::eigs_sym`.

### `RSpectra::eigs_sym` is used to calculate 10 smallest eigenvalues of $\boldsymbol{\Sigma}^{-1}$, and their inverse are corresponding to the top 10 eigenvalues of $\boldsymbol{\Sigma}$.
```{r}
ar_length <- 4096
auto_corr <- .9 # Corresponds to `\phi` above

ar_precision_matvec <- function(v, auto_corr) {
  # Fill in: note that you can vectorize the calculation and do *not* need a for-loop. 
  # (Hint: how would you efficiently carry out a matrix-vector operation if the matrix has non-zero entries only along the sub or super diagonal?)
  p<- auto_corr
  n_v <- length(v)
  v_1n<-c((cbind(v[1:(n_v-2)],v[2:(n_v-1)], v[3:n_v])%*%c(-p,1+p^2,-p)))
  1/(1-p^2)*c(v[1]-p*v[2],v_1n,-p*v[n_v-1]+v[n_v])
}

ar_eig <- eigs_sym(
  A = ar_precision_matvec, args = auto_corr,
  # Fill in
  n = ar_length,
  k=10,
  which = 'SM',
  opts = list(
    ncv = 100, # Spectrum distribution of AR-1 process is not very spread out on the extreme ends and is actually a hard case for Lanczos. So it helps to have more Lanczos vectors than the default for faster convergence.
    maxitr = 10^3, # Cap it just in case
    retvec = TRUE # More efficient to do without eigenvectors when not needed
  ),
)
```

Now, directly compute the eigen decomposition of $\boldsymbol{\Sigma}$ (not $\boldsymbol{\Sigma}^{-1}$) and compare its output with the principal components and associated variances found via Lancsoz algorithm.

### The direct_method function can be used to calculate the cov matrix for AR-1. For the comparison, we can use the results are close. 
```{r}
# Fill in
direct_method<-function(n = ar_length,p = auto_corr,k = 10){
  sig<-sapply(1:n, function(i){
    p^abs(i-c(1:n))
  })
  res<- eigen(sig)
  res_vec<-res$vectors[,1:k]
  res_val <- res$values[1:k]
  return(list(val = res_val, vec = res_vec))
}
res<-direct_method()
sum((res$val - 1/ar_eig$values[10:1])^2)
sum((abs(res$vec) - abs(ar_eig$vectors[,10:1]))^2)
```

**Remark:** 
For banded matrices, there actually are even more efficient approaches.
To get a sense of special routines available for banded matrices, you can take a look at `*_banded` functions in [SciPy's linear algebra routines](https://docs.scipy.org/doc/scipy/reference/linalg.html).
Even those functions represent only a subset of available numerical linear algebra techniques; see
[LAPACK documentation for SVD](https://www.netlib.org/lapack/lug/node32.html) for example.
